Primate Cognition and the Red Queen Hypothesis

### 2021-05-25 Article/Article.Rmd
Article modified after Benoit's comments

output: 
 #      - '--lua-filter=scholarly-metadata.lua'
 #      - '--lua-filter=author-info-blocks.lua'
 #bookdown::pdf_book:
 bookdown::word_document2:
   number_sections: false
   toc: false
   #citation_package: natbib
   latex_engine: pdflatex
   fig_caption: true
   pandoc_args:
     - '--lua-filter=scholarly-metadata.lua'
     - '--lua-filter=author-info-blocks.lua'

@@

**Abstract:** The diversity in animal cognition raises the question of its underlying evolutionary drivers. Cognitive selection can stem from interactions of individuals with their environment (*Ecological Intelligence Hypothesis*), with conspecifics within the social unit (*Social Intelligence Hypothesis*) or between social units (*Napoleonic Intelligence Hypothesis*), or even among generations (*Cultural Intelligence Hypothesis*). Yet, one species rarely occupies an area alone: space is a shared place between many species that can interact directly or indirectly. These interactions could represent additional cues and/or noise to find resource or escape predator(s), henceforth stimulate cognition. To test whether species co-occurrence shaped current patterns of cognition, we used Primates (N$_{min-max}$=`r minAllAreas`-`r maxAllAreas`) as a study example and retraced the evolutionary history of species whole brain - or specific area – relative size, using standard phylogenetic models or competitive phylogenetic models. Then we fitted phylogenetic regressions to test for variation in brain size and current co-occurrence pattern correlations. We found that the evolution of the relative size of areas involved in foraging, information processing and/or retention is better described by models accounting for species co-occurrence within dietary guilds which lead to increased relative sizes of these areas. This suggests that cognition is subject to an arms race among species (*Red Queen Intelligence Hypothesis*) and shall explain why specific abilities particularly boomed for some lineages, as for instance with Primates.

@@

To grasp the rationale of current hypotheses that aim to describe the evolution of animal cognition [@van2006some; @dunbar2017there], consider an animal foraging for food. This focal is not looking for any type of food but specific one. This can for instance be fruits. Finding fruits for an animal is not an easy task. In fact, fruits are the archetype of a hard-to-find resource likely to promote cognitive development because fruit trees are rare, dispersed, and do not produce constantly along the year nor between years, albeit their fruiting period remains nonetheless predictable [@janmaat2016spatio]. Additionally, they are energy-rich, allowing expansion of costly tissues such as brain tissues [@clutton1980primates]. Where the example focal stands there is no food that is immediately perceived, and the focal must move beyond its sensory range if it is to feed. Moving at random could mean travelling unnecessary long distances to finally rich a potentially void tree. Hence, for the focal, being capable of processing immediate cues and clues to know where specifically to find fruit trees [@dall2005information; @grove2013evolution] and whether it is probable that these trees currently yield fruits [@dall2005information; @robirainreview; @janmaat2016spatio] would be a considerable advantage. Provided it has the appropriate cognitive machinery, this could lead to avoiding costly detours and time recursions to food patches adequately. This simple picture draws the basis of the *Ecological Intelligence Hypothesis* [@clutton1980primates; @milton1981distribution; @rosati2017foraging] which stipulates that cognition was selected as a way to buffer the spatio-temporal complexity of the environment (e.g. @van2012large). In its quest, however, the focal might not be alone: there might be a few conspecifics foraging on its side, each looking for a signal leading to food sources. With these conspecifics, the focal forms a social group: they spend considerable time together, have established bounds and constantly share information. It is thus collectively that they must find a fruiting tree. To do so, being able to process signals emitted by others would be an advantage, as well as reading through their mind [@devaine2017reading] in order to plan ahead for collective actions or even outsmart them [@byrne1994machiavellian]. This therefore means plenty of additional stimuli to process, and constitutes the socle of the *Social Intelligence Hypothesis* [@dunbar2017there]. In the focal group, however, all individuals are not equally knowledgeable: perhaps because some have more experience than others do, such as the oldest would be more experienced than the youths. It is possible that these young individuals learn from the elderly individuals [@laland2004social]. For instance, they may learn how to process a specific tool to access a hidden comestible part as for nuts [@boesch1994nut]. In other words, there is vertical transmission that passes by over generations and knowledge thereby accumulates: this is the *Cultural Intelligence Hypothesis* [@whiten2007evolution, reader2002social], also known as the *Vygotskian Intelligence Hypothesis* in humans [@herrmann2007humans; @tomasello2009cultural; @wilson1991molecular; @van2011social]. Finally, while the focal’s group peacefully forages, it might sense the past or current presence of another group and anticipate its move, such as wild baboons do when prioritizing food likely to be depleted by other troops [@noser2010wild]. The neighbouring groups might indeed be competitors for food, mates or actually any resource that it might be essential to preserve, and as much as it is advantageous to be able to read through the environment and intra-group conspecifics, it might be advantageous to be able to decode information relative to other groups’ presence and attribute (e.g. group size), a view brought by the *Napoleonic Intelligence Hypothesis* [@ashton2020interactions].

|   In this overly simplistic picture, we moved from a unique individual to a group of individuals and then to multiple groups. Never was it question of multiple species: all hypothesized individuals belonged to a singular common species. Yet, a species rarely occupies a given area alone. Space is a shared place between a plethora of species, some of them occupying a same ecological niche. This partial niche overlap could contribute to impoverishing food density and increasing its spatio-temporal variance, hence reinforcing the selective pressure towards advanced cognition to maintain sufficient foraging success. In another perspective, two species sharing common food targets might benefit from being able to process cues of passage or presence of the other, to infer where and when to find ripe resource. To sum up, species co-occurrence might contribute into shaping animal cognition: the more species, the more advanced the cognition because the more noise or stimuli. In this study, we therefore propose to test for the existence of such an arms race [@van1973new] in brain size among species, what we naturally term the *Red Queen Intelligence Hypothesis*, building on frugivorous primates as a study example.

@@

|   Recent tools have been developed to infer the effect of species interaction on trait evolution, either by modelling divergence of the trait in co-occurring species from a same guild (e.g. dietary guild) (Matching Competition: MC models), or due to a density dependence of the evolutionary cladogenesis rate (linear: DD$_{lin}$ or exponential: DD$_{exp}$; @drury2016estimating) on the number of lineages within the guild. After reconstructing the range history based on `r minRange` to `r maxRange` primate species (@matzke2013probabilistic; @matzke2016stochastic; Figure \@ref(fig:figmap)) as well as primate diet evolution based on `r minFruit + minLeaf` to `r maxFruit + maxLeaf` species (discrete trait: frugivory vs. folivory; @bollback2006simmap), we calculated the likelihoods of models considering the role of species interactions (including competitive scenarios) in the evolution of either the whole brain (using the encephalic quotient, $EQ$, as a proxy for `r minEQ` to `r maxEQ` frugivorous, depending on how frugivory was assessed, see Dietary guild), or the relative size of specific brain areas associated with foraging-related information perception, processing or retention (Figure. \@ref(fig:figbrain); comprising `r minAllAreas` to `r max(maxNeocortex,maxHippocampus,maxCerebellum,maxStriatum,maxMOB)` frugivorous species) as well as the likelihoods of simpler models assuming non effect of species interactions, like the pure drift; Brownian Motion (BM), the Ornstein-Uhlenbeck process (OU) assuming that traits are constrained around on optimal value (e.g. stabilizing selection; see @blomberg2020beyond for a review on these approaches) or the Early-Burst model (EB, REF), this latter allowing to check for a time-dependence of the evolutionary rate, hence emphasizing that, if any, the density effect is not an artefact due to time dependence. Support for each model was evaluated using an information-theoretic framework [@burnham2002model] based on the weight of Akaike Information Criterion (AIC) corrected for small samples (AICc) when considering all five models (MC, DD$_{lin}$, DD$_{exp}$, BM, OU, EB, see Phylogenetic models; Evolutionary models in methods). Non-competitive models were the most likely in describing the evolutionary history of the $EQ$ or the striatum (Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)), an area specifically involved in XXX, used here as a control area. By contrast, competitive models were most supported in areas involved in sensory abilities (the main olfactory bulb, MOB), immediate information processing (i.e. working memory, with the neocortex), or long-term spatio-temporal information retention (hippocampus), (Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)). When density-dependent models were the best fit, the positive rate (b or r, Figure \@ref(fig:figresultsevolution)) suggested an acceleration of the evolutionary tempo together with increased lineage diversity.

|   To understand the directionality of the selection gradient shaped by co-occurrence (i.e. selection for “bigger” brain or “smaller” brain the more species), we fitted phylogenetic regressions (see Phylogenetic models; phylogenetic regressions in methods). In these linear regressions, the predicted variable was the relative brain size values of areas for which we observed interspecific competition effect. we considered the average surface of the frugivorous species range overlapped by other sympatric frugivorous species, as well as the number of such sympatric frugivorous species across their entire range as covariates.  Despite considerable noise induced by recent increase extinction and spatial range reduction due to anthropogenic disturbances (REF), it shows that XXXXXXXXXX

@@

Data processing, analyses, and plots were computed with R software version 4.0.3 [@Rsoftware]. Used datasets and codes will be made freely available upon publication. They can also be provided during the reviewing process for verification of missed coding errors, or any further investigation by the reviewers.

@@

We used a block of chronogram trees of the Primates taxon of the 10kTrees project (downloaded on the 11/05/2021, version 3), as well as a consensus tree of 1 000 trees for the subsequent phylogenetic analyses. The trees contain `r length(phylo_init$tip.label)` primate species.

@@

Brain data were obtained from @decasien2019primate for whole brain and all mentioned other parts (cerebellum, hippocampus, main olfactory bulb (MOB), neocortex, striatum) , @powell2017re and @powell2019maternal for whole brain, cerebellum and neocortex size, @todorov2019primate for hippocampus and neocortex size, @grueter2015home for the whole brain size and @navarrete2018primate for the whole brain, cerebellum, hippocampus and striatum size. They were freely available in the main manuscript or supplementary materials. If the species was represented multiple times in the dataset, we obtained a unique attribute by averaging it. From the global endocranial brain volume, we obtained the Encephalization Quotient (*EQ*, (N$_{max}$=`r maxEQ`)) as follows [@decasien2017primate]

@@

One to multiple large-scale geographic areas were assigned to each species as soon as the species current range overlapped in surface at `r geographicThresholdVector[1]` (low threshold) or `r geographicThresholdVector[2]`% (high threshold; the maximum was chosen to `r geographicThresholdVector[2]`% because on present data, a species could occupy as far as three areas) with the individual geographic area (Figure \@ref(fig:figmap); “gIntersection” function from the *rgeos* package [@rgeos] applied to Mercator-projected data to obtain the overlapping area, “area” function from the *geosphere* package [@geosphere], applied directly on unprojected longitudinal-latitudinal data for area calculation). These geographic areas were manually delimited using Google earth pro (version X.X, REF) as a combination of the environment topology and geographic regionalization relative to the primate taxonomy [@kamilar2009environmental]. Based on the structure (i.e. number of species and their phylogenetic relationship) of primate communities at different field sites, @kamilar2009environmental determined clusters of sites with highly similar community structures that were shaped by both the environment geography and climatic correlates. The considered geographic areas are represented in Figure \@ref(fig:figmap). The chosen scale for the areas is large because (1) retracing history of a large number of areas necessitates considerable computational means. In addition, this drastically increases computational time of phylogenetic model of brain trait evolution too. Furthermore (2), all species and particularly primate species particularly suffer from recent extinction [@pavoine2019mammal], with reduction of ranging areas at an unpreceeding speed rate. Finer geographic characterization would therefore give too much weight to this recent anthropogenic effect that recently altered species distribution (e.g. evidenced on the North American fauna in @pineda2021mammal).  Finally, note that the north part of Africa and the south of Europe were discarded despite the presence of one primate species (*Macaca sylvanus*), because of its geographical complete isolation and repeated intervention of human people in population maintenance [@modolo2005phylogeography]. Hence, *Macaca Sylvanus* is not considered in this study.
	We retraced the history of the lineage ranges based on current observation of species range using the *BioGeoBEARS* package [@matzke2013probabilistic] following the biogeographic stochastic mapping algorithm [@matzke2016stochastic]. This algorithm aims to fit, among others, non-time-stratified dispersal-extinction-cladogenesis (DEC) models (used here), specifically suiting analysis of ranging data since it accounts for spatially explicit processes of cladogenetic and anagenetic events (see @matzke2013probabilistic for further details on these events). To reconstruct the evolution of species range, WE fixed the maximum numbers of areas that could be occupied by a lineage at one time to three areas. A too high number of areas that can be occupied simultaneously drastically increases computational time. Here, we therefore chose three areas since it offers the possibility to occupy a complete mainland continent. Finally, because these history reconstructions are likely to vary, for each run of DEC models, we obtained `r numberSimulations` stochastic maps that were all used in subsequent phylogenetic model fitting (Phylogenetic models), such as it allowed me to account for uncertainty of these ancestral range estimations (see Phylogenetic models, Evolutionary Models (b)).

@@

We classified species as either frugivorous, folivorous or “other” based on the availability of frugivorous rate and folivorous rate, prioritizing fruvigory over folivory. First, a species would be classified as frugivorous if the frugivory rate was at least above `r frugivoryThresholdVector[1]` (low threshold) or `r frugivoryThresholdVector[2]`% (high threshold). If this was not the case, or frugivory rate was unavailable, a species could be classified as folivorous if the folivory rate was at least above `r folivoryThresholdVector[1]` (low threshold) or `r folivoryThresholdVector[2]`% (high threshold). Otherwise, @decasien2017primate gave a binary classification of diet, species being categorized as frugivorous or folivorous, partly based on anatomical criteria. Whenever the rate was not available, we referred to this classification. In any other cases, the species was classified as “other”.
	Frugivory rate was prioritized over, folivory, because we considered that since fruits are a highly palatable food source, it would be the key item that drives the foraging strategy (and associate consequence on brain selection), even if less consumed. Additionally, to consider frugivory, we used a lower rate than for folivory for two reasons. First, such static rate does not reflect potential seasonality in fruit eating [e.g. @masi2009western], which is generally shorter, hence a lower overall frugivory rate. Second, frugivory rate is likely to be underestimated in part because primates generally spend more time eating leaves than fruits, while rates are often based on eating relative feeding time, or observation frequency at the individual or group unit of feeding event. Finally, because the methodology to obtain this rate could additionally vary (e.g. in addition to the two aforementioned estimations, one could also rely on the proportion of species targeted for their fruits/leaves), We used two threshold levels (low, `r frugivoryThresholdVector[1]`%, or high, `r frugivoryThresholdVector[2]`%) to classify a species as frugivorous, as well as two threshold levels (low, `r folivoryThresholdVector[1]`%, or high, `r folivoryThresholdVector[2]`%) to classify a species as folivorous.
Considering diet as a categorical variable (frugivory vs folivory), we retraced the evolutionary history of such discrete traits based on a continuous Markovian process (extended Mk models) and relying on a Bayesian approach [@bollback2006simmap], using the “simmap” function of the *phytools* package [@phytools] and internally estimating the prior probability of trait (i.e. at the root) but with no prior on the transition matrix. Again, the obtained character history is in no case certain. Therefore, for each run, we obtained `r numberSimulations` stochastic character maps that were used in subsequent phylogenetic model fitting (Phylogenetic models), such as it allowed me to account for uncertainty of these ancestral diet estimations (see Phylogenetic models, Evolutionary Models (b)).

@@

We focused on frugivorous primates, because sample size was otherwise insufficient, and fitted phylogenetic model of $EQ$ - or relative size of a specific brain area – evolution with and without species competitions. Models were fitted on different sample sizes due to non availability of some data. Specifically, models using $EQ$ included `r minEQ` to `r maxEQ` frugivorous species. Other models included more reduced sample sizes (in species number): striatum (`r minStriatum` to `r maxStriatum`), MOB (`r minMOB` to `r maxMOB`), neocortex (`r minNeocortex` to `r maxNeocortex`), hippocampus (`r minHippocampus` to `r maxHippocampus`), cerebellum (`r minCerebellum` to `r maxCerebellum`). Prior fitting, trait parameters were log-transformed in case the distribution was skewed (i.e. for brain and MOB ratio) to reach more symmetrical distributions. Models without competition, Brownian motion (i.e. BM, pure drift model), Orstein-Uhlenbeck process (i.e. OU, model with stabilizing selection), or Early-Burst model (i.e. EB, for assessing a time-dependence of the evolutionary rate) were fitted using the “fitContinuous” function from the *geiger* package [@geiger3; @geiger5]. Using the evolutionary history of species distribution (Primate co-occurrence paragraph) and of diet (Dietary guild paragraph), we fitted competitive models using the “fit_t_comp” function from the *RPANDA* package [@RPANDA]. These competitive models notably account for interaction matrices that are build on the evolutionary history of species co-occurrence and diet. These interaction matrices retrace, along the phylogenetic tree, which lineages (species or ancestors) of a same dietary guild were present within the same geographic areas (see @drury2016estimating). We fitted three different competitive models. The matching competition model (MC) considers divergence of traits of co-occurring lineages from a same dietary guild due to repulsion of traits [@drury2016estimating]. Here, that would mean that co-occurring species would tend to have either extremely low or high $EQ$ or relative brain size. Otherwise, we modelled trait evolution accounting for linear (DD$_{lin}$) or exponential (DD$_{exp}$) density-dependence [@drury2016estimating; @weir2013diversity]. Density-dependence means that the evolutionary rate $\lambda$ varies either positively or negatively as a function f of the number of co-occurring lineages sharing the same diet such as

@@

(a)	Model dataset and formula

@@

(b)	Model implementation

@@

(c)	Model robustness

>>>>

output: 
 #      - '--lua-filter=scholarly-metadata.lua'
 #      - '--lua-filter=author-info-blocks.lua'
 bookdown::pdf_book:
 #bookdown::word_document2:
   number_sections: false
   toc: false
   #citation_package: natbib
   latex_engine: pdflatex
   fig_caption: true
   pandoc_args:
     - '--lua-filter=scholarly-metadata.lua'
     - '--lua-filter=author-info-blocks.lua'

@@

**Abstract:** The diversity in animal cognition raises the question of its underlying evolutionary drivers. Selection upon more advanced cognitive abilities can stem from interactions of individuals with conspecifics within the social unit (*Social Intelligence Hypothesis*), among generations (*Cultural Intelligence Hypothesis*), between social units (*Napoleonic Intelligence Hypothesis*), or with the rest of their environment (*Ecological Intelligence Hypothesis*). These hypotheses were limited to within-species scenarios. Yet, one species rarely occupies an area alone: space is a place shared between many species that can interact directly or indirectly. For species occupying a same ecological niche, interspecific competition is particularly expected in this case, but interactions could also represent additional cues and/or noise to cognitively process to find resource or escape predator(s) more efficiently. To test whether species co-occurrence shaped current patterns of cognition, we used Primates as a study example and retraced the evolutionary history and diversification of multiple species brain areas involved in foraging activities or not while considering competitive or non-competitive evolutionary scenarios. We found that the evolution of the relative size of areas involved in foraging-related information processing and/or retention is better described by models accounting for species co-occurrence within dietary guilds which lead to increased relative sizes of these areas. This suggests that cognition is subject to an arms race among species (*Red Queen Intelligence Hypothesis*) and shall explain why specific abilities particularly boomed for some lineages, as for instance with Primates.

@@

To grasp the rationale of current hypotheses that aim to describe the evolution of animal cognition [@van2006some; @dunbar2017there], consider a primate individual foraging for food. Primates are pivotal species to study the evolution of cognition given the complexity of their socio-ecological environment and the inevitable implication for assessing human evolution [@byrne2000evolution]. The chosen focal is not looking for any type of food but specific one, as for instance, most primate species often look for fruits. Finding fruits for a primate is not an easy task. In fact, fruits are the archetype of a hard-to-find resource likely to promote cognitive development because fruit trees are rare, dispersed, and do not produce constantly along the year nor between years, albeit their fruiting period remains nonetheless predictable [@janmaat2016spatio]. Additionally, they are energy-rich, allowing expansion of costly tissues such as brain tissues [@clutton1980primates].  Overall, primates thus show remarkable abilities to navigate precisely and target fruit trees likely to yield ripe resource despite they are out of their sensory range [@trapanese2019review], in part because of advanced spatio-temporal information retention [e.g. spatial knowledge: @normand2009forest; @robiraroute, temporal knowledge: @janmaat2006evidence; @janmaat2013chimpanzees; @janmaat2013tai].

|	When the resource is out of sight, moving at random could mean travelling unnecessary long distances to finally reach a potentially void tree. Hence, for the focal, being capable of processing immediate cues and clues to know where specifically to find fruit trees [@dall2005information; @grove2013evolution] and whether it is probable that these trees currently yield fruits [@dall2005information; @robirainreview; @janmaat2016spatio] would be a considerable advantage. Having appropriate cognitive machinery may avoid costly detours and time recursions to food patches adequately. This simple picture draws the basis of the *Ecological Intelligence Hypothesis* [@clutton1980primates; @milton1981distribution; @rosati2017foraging] which stipulates that cognition was selected as a way to buffer the spatio-temporal complexity of the environment (e.g. @van2012large).

|	In its quest for food, however, the focal might not be alone: for instance, primates often form social groups [@kappeler2002evolution] where group members spend considerable time together, have established bounds and constantly share information. Thus, they might altogether look for, and process, cues and clues leading to food sources. To do so, being able to process signals emitted by other maneuvers would be an advantage, as well as reading through their mind [@devaine2017reading] in order to plan ahead for collective actions or even outsmart them [@byrne1994machiavellian]. This therefore means plenty of additional stimuli to process, and constitutes the socle of the *Social Intelligence Hypothesis* [@dunbar2017there].

|   In the focal group, however, all individuals are not equally knowledgeable: perhaps because some have more experience than others do, such as the oldest would be more experienced than the youths. It is possible that these young individuals learn from the elderly ones [@laland2004social]. For instance, they may learn how to process a specific tool to access a hidden comestible part as for nuts [@boesch1994nut]. In other words, there is vertical transmission that passes by over generations and knowledge thereby accumulates: this is the *Cultural Intelligence Hypothesis* [@whiten2007evolution, reader2002social], also known as the *Vygotskian Intelligence Hypothesis* in humans [@herrmann2007humans; @tomasello2009cultural; @wilson1991molecular; @van2011social].

|   Finally, while the group peacefully forages, it might sense the past or current presence of another group and anticipate its move, such as wild baboons do when prioritizing food likely to be depleted by other troops [@noser2010wild]. The neighboring groups might indeed be competitors for food, mates or any essential resource, and as much as it is advantageous to be able to read through the environment and intra-group companions, it might be advantageous to be able to decode information relative to other groups’ presence and attribute (e.g. group size), a view brought by the *Napoleonic Intelligence Hypothesis* [@ashton2020interactions].

|   In this overly simplistic picture, we moved from a unique individual to a group of individuals and then to multiple groups. Never was it question of multiple species: all hypothesized individuals belonged to a singular common species. Yet, space is a place shared between a plethora of species, some of them occupying a same ecological niche. As much as conspecifics could be competitors, or direct or indirect cooperators, so could be individuals from an other species with similar ecological preferences, for instance with regards to diet. As such, co-occurrence of species from the same guild might contribute into shaping animal cognition: the more species, the more selective pressure for advanced the cognition . In this study, we therefore propose to test for the existence of such an arms race [@van1973new] in cognitive abilities among species - approximated by the size of the whole or specific regions of the brain - that we naturally term the *Red Queen Intelligence Hypothesis*, building on frugivorous primates as a study example.

@@

|   Recent tools have been developed to infer the effect of species interactions on trait evolution, either by modelling trait divergence in co-occurring species from a same guild (e.g. dietary guild) (Matching Competition: MC models) or considering that the evolutionary rate depends on the on the number of lineages within the guild (density dependence; linear: DD$_{lin}$ or exponential: DD$_{exp}$; @drury2016estimating). After reconstructing the biogeography history based on `r minRange` to `r maxRange` primate species (@matzke2013probabilistic; @matzke2016stochastic; Figure \@ref(fig:figmap)) as well as primate diet evolution based on `r minFruit + minLeaf` to `r maxFruit + maxLeaf` species (discrete trait: frugivory vs. folivory; @bollback2006simmap), we calculated the likelihoods of models considering the role of species interactions (including competitive scenarios) in the evolution of either the whole brain (using the encephalic quotient, $EQ$, as a proxy for `r minEQ` to `r maxEQ` frugivorous, depending on how frugivory was assessed, see Dietary guild), or the relative size of specific brain areas associated with foraging-related information perception, processing or retention (Figure. \@ref(fig:figbrain); comprising `r minAllAreas` to `r max(maxNeocortex,maxHippocampus,maxCerebellum,maxStriatum,maxMOB)` frugivorous species) as well as the likelihoods of simpler models assuming non effect of species interactions, like the simple Brownian Motion (BM), the Ornstein-Uhlenbeck process (OU) assuming that traits are constrained around on optimal value (e.g. stabilizing selection; see @blomberg2020beyond for a review on these approaches) or the Early-Burst model (EB, REF), this latter allowing to check for a time-dependence of the evolutionary rate, hence emphasizing that, if any, the density effect is not an artefact due to time dependence. Support for each model was evaluated using an information-theoretic framework [@burnham2002model] based on the weight of Akaike Information Criterion corrected for small samples (AICc) when considering all six models (MC, DD$_{lin}$, DD$_{exp}$, BM, OU, EB, see Phylogenetic models; Evolutionary models in methods). Non-competitive models were the most likely in describing the evolutionary history of the $EQ$ or the striatum (Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)), an area specifically involved in XXX, used here as a control area. By contrast, competitive models were most supported in areas involved in sensory abilities (the main olfactory bulb, MOB), immediate information processing (i.e. working memory, with the neocortex), or long-term spatio-temporal information retention (hippocampus), (Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)). When density-dependent models were the best fit, the positive rate (r, Figure \@ref(fig:figresultsevolution)) suggested an acceleration of the evolutionary tempo together with increased lineage diversity.

|   To understand the directionality of the selection gradient shaped by co-occurrence (i.e. selection for “bigger” brain or “smaller” brain the more species), we fitted phylogenetic regressions (see Phylogenetic models; phylogenetic regressions in methods). In these linear regressions, the predicted variable was the relative brain size values of areas for which we observed interspecific competition effect. We considered the average surface of the frugivorous species range overlapped by other sympatric frugivorous species, as well as the number of such sympatric frugivorous species across their entire range as covariates.  Despite considerable noise induced by recent increase extinction and spatial range reduction due to anthropogenic disturbances (REF), it shows that XXXXXXXXXX

@@

Data processing, analyses, and plots were computed with R software version 4.0.3 [@Rsoftware]. Used datasets and codes will be made freely available upon publication. Used codes and data are freely available at XXXXX.

@@

We used a block of chronogram trees of the Primates taxon of the 10kTrees project (downloaded on the 11/05/2021, version 3), as well as a consensus tree of 1,000 trees for the subsequent phylogenetic analyses. The trees contain `r length(phylo_init$tip.label)` primate species.

@@

Brain data were obtained from @decasien2019primate for whole brain and all mentioned other parts (cerebellum, hippocampus, main olfactory bulb (MOB), neocortex, striatum) , @powell2017re and @powell2019maternal for whole brain, cerebellum and neocortex size, @todorov2019primate for hippocampus and neocortex size, @grueter2015home for the whole brain size and @navarrete2018primate for the whole brain, cerebellum, hippocampus and striatum size. They were freely available in the main manuscript or supplementary materials. When the species was represented multiple times in the dataset, we obtained a unique attribute by averaging it. From the global endocranial brain volume, we obtained the Encephalization Quotient (*EQ*, (N$_{max}$=`r maxEQ`)) as follows [@decasien2017primate]

@@

One to multiple large-scale geographic areas were assigned to each species as soon as the species current range overlapped in surface at `r geographicThresholdVector[1]` (low threshold) or `r geographicThresholdVector[2]`% (high threshold; the maximum was chosen to `r geographicThresholdVector[2]`% because on present data, a species could occupy as far as three areas) with the individual geographic area (Figure \@ref(fig:figmap); “gIntersection” function from the *rgeos* package [@rgeos] applied to Mercator-projected data to obtain the overlapping area, “area” function from the *geosphere* package [@geosphere], applied directly on unprojected longitudinal-latitudinal data for area calculation). These geographic areas were manually delimited using Google earth pro (version X.X, REF) as a combination of the environment topology and geographic regionalization relative to the primate taxonomy [@kamilar2009environmental]. Based on the structure (i.e. number of species and their phylogenetic relationship) of primate communities at different field sites, @kamilar2009environmental determined clusters of sites with highly similar community structures that were shaped by both the environment geography and climatic correlates. The considered geographic areas are represented in Figure \@ref(fig:figmap). The chosen scale for the areas is large because (1) retracing history of a large number of areas necessitates considerable computational means. In addition, this drastically increases computational time of phylogenetic model of brain trait evolution too. Furthermore (2), all species and particularly primate species particularly suffer from recent extinction [@pavoine2019mammal], with reduction of ranging areas at an unpreceeding speed rate. Finer geographic characterization would therefore give too much weight to this recent anthropogenic effect that recently altered species distribution (e.g. evidenced on the North American fauna in @pineda2021mammal).  Finally, note that the north part of Africa and the south of Europe were discarded despite the presence of one primate species (*Macaca sylvanus*), because of its geographical complete isolation and repeated intervention of human people in population maintenance [@modolo2005phylogeography]. Hence, *Macaca sylvanus* is not considered in this study.
	We retraced the history of the lineage ranges based on current observations of species range using the *BioGeoBEARS* package [@matzke2013probabilistic] following the biogeographic stochastic mapping algorithm [@matzke2016stochastic]. This algorithm aims to fit, among others, non-time-stratified dispersal-extinction-cladogenesis (DEC) models (used here), specifically suiting analyses of range data since it accounts for spatially explicit processes of cladogenetic and anagenetic events (see @matzke2013probabilistic for further details on these events). To reconstruct the evolution of species range, we fixed the maximum numbers of areas that could be occupied by a lineage at one time to three areas. A too high number of areas that can be occupied simultaneously drastically increases computational time. Here, we therefore chose that a species can at most occupy three areas since it offers the possibility to occupy a complete mainland continent. Finally, because these history reconstructions are likely to vary, for each run of DEC models, we obtained `r numberSimulations` stochastic maps that were all used in subsequent phylogenetic model fitting (Phylogenetic models) to account for uncertainty of these ancestral range estimations (see Phylogenetic models, Evolutionary Models (b)).

@@

We classified species as either “frugivorous” or “folivorous” based on the availability of frugivorous rate and folivorous rate, prioritizing fruvigory over folivory. First, a species would be classified as frugivorous if the frugivory rate was at least above `r frugivoryThresholdVector[1]` (low threshold) or `r frugivoryThresholdVector[2]`% (high threshold). If this was not the case, or frugivory rate was unavailable, a species could be classified as folivorous if the folivory rate was at least above `r folivoryThresholdVector[1]` (low threshold) or `r folivoryThresholdVector[2]`% (high threshold). Otherwise, @decasien2017primate gave a binary classification of diet, species being categorized as frugivorous or folivorous, partly based on anatomical criteria. Whenever the rate was not available, we referred to this classification. In any other cases, the species was discarded.
	Frugivory rate was prioritized over folivory because we considered that since fruits are a highly palatable food source, it would be the key item that drives the foraging strategy (and associate consequence on brain selection), even if less consumed. Additionally, to consider frugivory, we used a lower rate than for folivory for two reasons. First, such static rate does not reflect potential seasonality in fruit eating [e.g. @masi2009western], which is generally shorter, hence a lower overall frugivory rate. Second, frugivory rate is likely to be underestimated in part because primates generally spend more time eating leaves than fruits, while rates are often based on eating relative feeding time, or observation frequency at the individual or group unit of feeding event. Finally, because the methodology to obtain this rate could additionally vary (e.g. in addition to the two aforementioned estimations, one could also rely on the proportion of species targeted for their fruits/leaves). We used two threshold levels (low, `r frugivoryThresholdVector[1]`%, or high, `r frugivoryThresholdVector[2]`%) to classify a species as frugivorous, as well as two threshold levels (low, `r folivoryThresholdVector[1]`%, or high, `r folivoryThresholdVector[2]`%) to classify a species as folivorous.
Considering diet as a binary variable (frugivory versus folivory), we retraced the evolutionary history of such discrete traits based on a continuous Markovian process (extended Mk models) and relying on a Bayesian approach [@bollback2006simmap], using the “simmap” function of the *phytools* package [@phytools] and internally estimating the prior probability of trait (i.e. at the root) but with no prior on the transition matrix. Again, the obtained character history is in no case certain. Therefore, for each run, we obtained `r numberSimulations` stochastic character maps that were used in subsequent phylogenetic model fitting (Phylogenetic models) to account for uncertainty of these ancestral diet estimations (see Phylogenetic models, Evolutionary Models (b)).

@@

We focused on frugivorous primates, because sample size was otherwise insufficient, and fitted phylogenetic model of $EQ$ - or relative size of a specific brain area – evolution with and without species competitions. Models were fitted on different sample sizes due to non-availability of some data. Specifically, models using $EQ$ included `r minEQ` to `r maxEQ` frugivorous species. Other models included more reduced sample sizes (in species number): striatum (`r minStriatum` to `r maxStriatum`), MOB (`r minMOB` to `r maxMOB`), neocortex (`r minNeocortex` to `r maxNeocortex`), hippocampus (`r minHippocampus` to `r maxHippocampus`), cerebellum (`r minCerebellum` to `r maxCerebellum`). Prior fitting, trait parameters were log-transformed in case the distribution was skewed (i.e. for brain, EQ, and MOB ratio) to reach more symmetrical distributions. Models without competition, Brownian motion (i.e. BM), Orstein-Uhlenbeck process (i.e. OU, model with stabilizing selection), or Early-Burst model (i.e. EB, for assessing a time-dependence of the evolutionary rate) were fitted using the “fitContinuous” function from the *geiger* package [@geiger3; @geiger5]. Using the evolutionary history of species distribution (Primate co-occurrence paragraph) and of diet (Dietary guild paragraph), we fitted competitive models using the “fit_t_comp” function from the *RPANDA* package [@RPANDA]. These competitive models notably account for interaction matrices that are built on the evolutionary history of species co-occurrence and diet. These interaction matrices retrace, along the phylogenetic tree, which frugivorous lineages were present within the same geographic areas (see @drury2016estimating). We fitted three different competitive models. The matching competition model (MC) may consider divergence of traits of co-occurring lineages from a same dietary guild due to repulsion of traits (character displacement) [@drury2016estimating]. Here, that would mean that co-occurring species would tend to have either lower or higher $EQ$ or relative brain size. Otherwise, we modelled trait evolution accounting for linear (DD$_{lin}$) or exponential (DD$_{exp}$) density-dependence [@drury2016estimating; @weir2013diversity]. Density-dependence means that the evolutionary rate $\lambda$ varies either positively or negatively as a function f of the number of co-occurring lineages sharing the same diet such as

@@

(a)	Determining the direction of the selection gradient shaped by interspecific competition

@@

(b)	Diversification analysis

TO BE COMPLETED

(c)	Model implementation

@@

(d)	Model robustness
